draft 1

### 2016-01-01 finalwriteup/draft.rmd
finished lit review

### 2.2.3 Introduction

### 2.2.4 Linear Regression/logistic regression

### 2.2.5 Naiive Bayes

### 2.2.6 Ensemble Trees

### 2.2.7 Bayesian Networks

### 2.2.7 SVM's and Neural Networks

### 2.2.8 Summary

@@

### 2.3.1 Most simple and frequent models

### 2.3.2 Complex models

* SVM, Neural networks, random forest
* data must not be normal
* most successful?

### 2.3.3 ensemble methods

* papers
* netflix?
* multi step to deal with missing data

### 2.3.4 similar applications

* Employee churn - anything similar?
* stepped method business application?
* Gap

**

@@

summary of 
* how are B2B decisions currently made
* what are the most popular/effective methods
* which ensemble methods may suit our missing data situation
* Gap in literature/contribution

>>>>

### 2.2.1 Introduction

The aim of this project is to use statistical techniques to model the profitability of projects. In order to optimise the model, several statistical and machine learning techniques will be tested, and compared and it is therefore prudent to review the range of statistical and machine learning techniques that have been successfully applied in the literature. These vary from simple methods such as Linear Regression to complex, deep learning such as Neural Networks. The previous section highlighted the use of linear regression, neural networks, SVM and in one case boosted trees for prediction, however other business problems have utilised a wider range of methods. These include Naiive Bayes, Random Forests, and machine learned Bayesian Networks. The following section will present each method's advantages and disadvatanges and an example of a past successful application.

### 2.2.2 Linear Regression

Linear regression is one of the simplest and most popular statistical prediction methods [@Hastie2009]. An equation for a line is learned, which is defined by a constant and a coefficient times the value of each variable. The coefficients are chosen to minimise the residual sum of squares (least squared) [@Hastie2009]. The reason this method is popular is that it is simple, provide stable predictions, and can be adapted for categorical variables [@Seng2010]. The integration of categorical variables is called Anova, and the algorithm compares group mean variances within the categorical variable.

There are several drawbacks to linear regression. For example assumptions must be made about the structure of the data. Predictor varialbes are assumed to be linearly related to the response variable and the varialbes are normally distributed. Many real world phenomena do not correspond to these assumptions which can make it difficult, sometimes impossible, to produce useful results [@Hastie2009][@Putler2012]. Furthermore, response variables must be continous numeric variables and if two or more predictor variables are correlated, the values of their coefficients are highly unstable. On the other hand, correlated variables do not impact the predictive accuracy of the model. A final drawback is that the variable coefficients cannot be used to compare variable importance unless the varialbes have all been standardised to one another [@Putler2012]. This limits the amount of insight that can be drawn from a linear regression model.

Despite the limitations of linear regression models, they have widespread business applications due to their simplicity and ease of application. Regression is used to predict risk in the finance and insurance industries, predict who to target in marketing exercises, and predict consumption spending in the field of economics [@Harrell2013]. Due to its simplicity and widespread application, linear regression is a good starting point for an analysis and benchmark for performance against complex models.

### 2.2.3 Logistic Regression

The method behind linear regression can be adapted to predict a binary response variable (1 or 0). A predictive linear equation is computed so that the predictor variables are combined with coefficients to give an output between 0 and 1. This is then transformed into the log odds to mediate extreme variable input values. The result is a probability an input case will result in a 0 or 1 (i.e. success or failure) [@Moore1989]. Similar pros and cons exist for logistic regression as linear regression, and it too can be used as a benchmark to compare other binary predictive models due to its speed and simplicity. Many business problems have binary response variables, such as yes/no, male/female, buy/do not buy, success/failure, or survive/death [@Moore1989].

### 2.2.4 Naive Bayes

Naive Bayes prediction method makes conditional independence assumptions about the predictor variables in order to greatly simplify probability calculations for the response variable (the response variable must be categorical). Specifically, it is assumed that each variable is conditionally independent given the respose class alone, without considering other predictor variables. Mathematically this means that each variable's probability contribution can be simply calculated by looking at the probability of a response class given a predictor value by itself. The predictor variables probability estimates are combined for each possible response class, and the response class with the highest probability is chosen [@Provost2013].

The advantages of this method are that the conditional independence assumption enables very fast calculations and predictions. The method can perform very well for real world tasks because the assumption of independence does not damage predictions significantly. This is because if multiple varialbes are related, the variables independently still direct the prediction in the correct direction. The correlated variables will double or triple their emphasis on the predicted response variable, however this does not necessarily impact classification accuracy as the final step is to simply choose the class with the *highest* probability [@Provost2013]. This is fine for ranking. On the other hand, the output probabilities are not realisitic, and the values themselves should not be used [@Caruana2006]. Another disadvantage is that in order to calculate the class probability from a numeric variable, a distribution must be assumed (often gaussian) which the numeric data may not fit into neatly [@Caruana2006]. Finally, Naive Bayes classifers are known to not perform well in binary classification [@Caruana2006].

A real world success story for Naive Bayes classifiers hinges on their ability to be refined incrementally with each piece of new data, as opposed to re-calibrating the entire model. Naive Bayes has been used in complex spam detection systems where new spam emails or toxic text themes can be quickly added to the filtering model. The Naive Bayes method provides a good benchmark to compare against more complex models that should outperform it [@Caruana2006].

### 2.2.5 Decision Trees/ Ensemble Trees

Decision trees are one of the simplest and most intuitive machine learning methods. There are several brands of basic decision tree algorithms including ID3, C4.5, CART, and CHAID with CART and C4.5 being the most popular [@Kabra2001]. Ravi Kumar and Ravi [@Kumar2007] recommend the CART algorithm as it is capable of solving both classification and regression problems whereas the remaining decision trees solve classification problems only. Single decision trees have the risk of over-fitting the data and low predictive accuracy but have favourable characteristics such as providing intuitive rules that a decision maker can follow in real-life scenarios. In addition their results do not depend heavily on the skill of the data analyst [@Kabra2001][@Putler2012].

To tackle the problem of low predictive accuracy and instability, ensemble tree methods were pioneered in the 1990's with success [@Breiman1996]. Three examples of these ensemble decision tree methods are bagging, boosting, and random forests. Bagging creates multiple trees by sampling a different training data set for each tree and then combining tree results [@Breiman1996]. Random forests is an advanced form of bagging in that multiple trees are sampled from the same training data. However, when creating the tree for each sample dataset, a random subset of attributes is used to determine each split. This prevents the trees from becoming too correlated [@Breiman2001]. The boosted decision tree approach combines gradient descent with a series of decision trees. Each tree is limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) from the preceding tree. The limited depth of each tree prevents overfitting at each stage and the combined result of up to thousands of trees is very powerful [@Elith2008].

Ensemble decision trees have a different foundation to more traditional statistical prediction methods such as regression. For this reason, there are some advantages and disadvantages. The progression of binary splits means non-linear feature interactions can be captured and the predictor variables do not need to be transformed as no assumptions are made about the data's statistical distributions [@Louppe2014][@Radenkovic2010]. Other advantages are that the ensemble methods are very fast in comparison to SVM's and neural networks but perform just as well and provide insights such as variable importance and variable relationships [@Sealfon2012]. Random forests in particular has the benefit of not being affected by noise in the data as well as the capacity to handle a large number of variables [@Sealfon2012]. Caruana [@Caruana2006] tested boosted trees, random forests, neural networks, SVM's, logistic regression and naive bayes on 11 binary classification problems and found that boosted trees performed best, followed by random forests. Clearly, ensemble tree methods can compete with high level machine learning algorithms.

Along with the many advantages of trees and ensemble trees, it is important to be aware of their limitations. As trees are not built on a probabilistic framework, results can not be provided in this framework. For example, confidence intervals for predictions are not available for standard ensemble methods [@Louppe2014]. Also, variable importance tables provided by random forests and boosted trees can be biased towards variables with many categories and another type of ensemble method, conditional forests, should be used [@Radenkovic2010]. The methods can also be prone to overfitting if not handled carefully. Ensemble trees have many advantages particularly predictive power, but it is important to recognise that the best method for a problem often depends on the type of problem [@Louppe2014][@Caruana2006].

### 2.2.5 Bayesian Networks

A bayesian network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a DAG (directed acyclic graph) and is capable of linking the conditional dependency between any variable to another variable or via other variables which are represetned by arcs [@Heckerman1998]. This kind of conditional relationship is bayesian. Bayesian networks have found success in combining deterministic models with observation data as well as expert knowledge as the statistical relationships between variables can be determined from different places and manually entered into a single model [@Kragt2009].

Bayesian networks are excellent at representing knowledge from different sources and types, and visuallising them in a way that communicates results well to decision makers [@Kragt2009]. Therefore bayesian networks become an excellent decision support tool. Network relatitonships can be machine learned, however this is not widely included in the suite of machine learning methods and is referred to more as a way to model complex networks. Drawbacks of bayesian networks include their inability to handle continuous numeric variables (which must be discretised). This means that the varialbe relationships are categorical [@Kragt2009]. Popular applications of bayesian networks are modelling uncertainty in natural resource management and modelling complex business network structures such as airports [@Wu2013].

### 2.2.6 SVM's and Neural Networks

Neural networks were inspired by biological neural networks of the human nervous system and require a lot of training data. They are very flexible in the relationships they can mimic and are not greatly dependent on the skills of the analyst (Putler & Krider, 2012). A disadvantage is that although the results are accurate, the algorithm is a black box - meaning the data is input and the results output without providing the user insight into the methodology. Neural networks can perform high level tasks as a result of deep learning such as hand writing recognition, vehicle control, face recognition, and cancer detection [@Haykin2004].

The simplest type of Support vector machines (SVM's) calculate the linear boundary between two categories in feature space so that the category of future cases may be determined depending on which side of the boundary they fall. The location of the boundary is determined by the position of the widest clear margin that can be drawn between the two categories, allowing a certain number of misclassifications. This concept can be adapted to non linear boundaries by performing kernel transformations on prior variables as well as regression problems by introducing loss functions [@Auria2008]. Advantages of SVM's include the ability to model non linear relationships in data, and that no assumptions about the distributions of the data must be made as the model is non-probabilistic. They are also known to be robust over different samples and perform well in high-dimensional feature space [@Auria2008]. On the other hand, the output is in the form of distance to the boundary (as opposed to probability) and results are not transparent (similar to neural networks) [@Caruana2006]. Also, SVM's are known to be very slow to train and therefore often not suitable for industry purposes [@Auria2008]. Because SVM's perform well in high-dimensional space and can form non-linear rules they have been applied to protein classification (medicine) and text and image recognition [@Byun2002].

### 2.2.7 Summary

This section reviewed the advantages and disadvantages of statistical and machine learning techniques that have been successfully applied to business problems. These include linear regression, logistic regression, naive bayes, decision trees and ensemble trees, bayesian networks, neural networks, and SVM's. Linear/logistic regression and naive bayes are computationally very fast with a simple concept that make broad assumptions about the data, but work well as baseline models against which to compare complex models. In certain cases these simpler models are significantly outperformed by complex models however it is important to compare both. Neural networks and SVM's are excellent machine learning predictors which can perform well with messy data and no assumptions, however they are slow and do not provide insight into the results of analysis. This is not appropriate for the profitability prediction problem in this project because communication of the reasoning behind predictions is crucial for uptake of the model by decision makers.

On the other hand, ensemble tree methods can perform as well as neural networks and SVM's and make no assumptions about the structure of the data, but provide insights into the model such as variable importance and variable relationships. Bayesian networks are excellent for presenting the structure and reasoning behind the predictive model, however are very slow to train and have generally not been included in the machine learning literature as much as the other methods. This may be due to low awareness of learned bayesian networks or lower performance. In conclusion, simple models such as linear/logistic regression and naive bayes are good baseline models to implement, while ensemble tree methods and bayesian networks may handle messy data better than the simpler models while still providing insight to decision makers.

@@

The aim of this research project is to create a predictive model for the profitability of a company's consulting projects using their internal CRM data. The literature pertaining closely to the cost prediction topic has been reviewed in section 2.1, however it is relevant to broaden the assessessment of the literature on advanced machine learning techniques that have been applied to general business problems. The simpler statistical methods have been covered already, so the focus of this sections is on the topic of machine learning prediction methods for business decision-makers. Popular applied topics include predicting stock fluctuation, customer churn analysis, fraud prediction, customer classification, market segment analysis, product success prediction, and recommendation systems (Seng & Chen, 2010). This section will first discuss the use and merits of the employed more advanced machine learning methods, followed by a case study of employee-churn prediction.

### 2.3.2 Summary of Advanced Statistical and Machine Learning Methods Applied to Business Problems

According to Breiman [@Breiman2001b], simple theoretical models such as linear regression are informative but often provide unsatisfactory accuracy for modelling real life data which is messy and does not necessarily align with the rules of statistical theory. As this project will address a problem with a large quantity of messy data, it is worthwhile reviewing the algorithmic machine-learning predictive models that have developed rapidly in recent years. They have found success because they are able to sometimes fit real-life data with more accuracy than theoretical models [@Breiman2001b]. Ravi Kumar and Ravi [@Kumar2007] performed a detailed review of statistical and machine learning techniques that were applied over 37 years in the context of bankruptcy prediction in banks. The most widely used model was neural networks, however logistic and linear regression, decision trees, SVM’s, discriminant analysis (DA), and statistical clustering techniques (such as K Nearest Neighbour) were also popular. Supporting Breiman’s [@Breiman2001b] statements, it was found that DA and linear regression techniques were not preferred due to their low accuracy. The overall assessment was that SVM’s outperformed Neural networks (back propagation neural networks were used most extensively), which sometimes outperformed decision trees, and the rest of the methods were generally inferior. Although SVM’s performed the best, as discussed, they are often extremely complex and slow, requiring a high memory [@Kumar2007]. In support of SVM predictive performance, a study by Davenport and Harris [@Davenport2007] concluded that statisticians experienced in predictive machine learning algorithms generally acknowledge that SVM’s yield the highest prediction accuracy compared to other machine learning algorithms.

As SVM’s are often computationally too expensive and advanced for a smaller scale business applications, such as this research project, it is useful to compare neural networks and decision trees which are simpler than SVM’s but are also capable of achieving very accurate predictive models. Ravi Kumar and Ravi [@Kumar2007] found that neural networks and decision trees were both capable of out-performing the other, depending on the context – which is limited to bankruptcy prediction in Kumar’s study. Both have advantages and disadvantages to consider (refer section 2.2), however the most notable drawback for neural networks is the difficulty in interpreting the algorithms' explanation of human behaviour – a valuable feature in business problems [@Tsai2009]. On the other hand, trees are known to be intuitive to understand and have been argued to function similarly to how the human mind thinks [@Carrizosa2010].

Ravi Kumar and Ravi's [@Kumar2007] study concludes that ensemble techniques, which refers to combinations of two completely different algorithms, can often outperform individual techniques. For example, combining the contrasting advantages of neural networks and decision trees is a worthwhile ensemble technique. A paper studying financial earnings management prediction combined these two methods by running them one after another. The neural network was run first to create a high rate of prediction accuracy via its complex non-linear learning properties [@Tsai2009].  Then, to resolve the lack of explanatory qualities that neural networks are notorious for, decision trees were employed. The 81% of cases that were correctly predicted by the neural network were used to generate decision trees, and in turn useful decision rules [@Tsai2009]. This strategy is a promising way to benefit from the strengths of well performing complementary techniques.

### 2.3.3 Employee Churn Case Study

The literature on predictive business models is generally limited to studies on product sales, customer behaviour, or overall performance of large businesses. There are far fewer studies that assist with internal business decisions dealing with employees, teams, and delivering projects as is the case with this project. An exception is Saradhi and Vijaya’s [@Saradhi2011] study on employee churn, where ‘churn’ refers to the number of individuals moving out of a group within a certain time. Saradhi and Vijaya applied popular customer churn predictive models to employee churn - a novel application that focussed internally on employees rather than externally on customers. The project chose three machine learning classification algorithms that are normally used in customer churn and it was proposed that the variables and model behaviour of employee churn could be related to customer churn. The associated costs of losing customers and finding new customers can also be correlated to the costs of losing staff and hiring new staff.

Naive Bayes, Random forests (an ensemble decision tree method) and SVM’s were built. A comparison study showed that Naïve Bayes is outperformed by random forests which match Saradhi and Vijaya’s prediction [@Caruana2006]. The SVM model actually far out performed random forests and Naive Bayes by achieving 81% vs 51% and 55% correct predictions respectively. This was attributed to the ability to introduce class penalties in SVMs whereas the other two methods are limited by class imbalance [@Saradhi2011]. This means that SVM’s had a superior method of balancing the importance of the variables in the data set which was skewed by a heavy majority of employees who kept their job versus the minority who left (25% churn). It should be noted that the boosted tree ensemble method performs a similar task of weighting misclassified cases or outliers and it would have been worthwhile comparing this method to SVMs. Particularly because boosted trees provide insight into the predictive model which could aid uptake of the model with decisoin makers. However, it is noteworthy that their study proved that predictive techniques used on customres could be translated to predicting internal behaviour of employees.

### 2.3.4 Gap/conclusion

The limited literature on analysis similar to Sardhi and Vijaya’s Employee Churn study [@Saradhi2011] highlights a gap in the application of predictive techniques to model internal performance in businesses. This is particularly relevant to consulting companies that tackle complicated discrete projects with set timelines. In this project, internal consulting project data will be used to predict a new project performance in terms of internal employee structure as well as client characteristics. A review of the literature in machine learning applied to business problems revealed that SVM's have found to be the most accurate technique followed by neural networks and ensemble tree methods. As discussed in section 2.2, it was also noted that neural networks and SVM's do not provide insight into the model's predictions to the decision maker. A gap exists in applying these cutting edge machine learning techniques to internal business decision-making.

@@

The literature pertaining to the aim of this research project, cost prediction using internal consulting CRM data, has been extensively reviewed. The problem of cost prediction has been addressed in many studies beforehand which have generally been limited to prediction of construction costs for buildings and infrastructure and effort estimation for IT projects. The prediction models consisted of case based reasoning, linear regression, neural networks and in one or two instances an ensemble tree or SVM. This leaves a gap in the literature in effort estimation for consulting companies in the construction industry as well as trialling a broader range of prediction methods.

An exhaustive list of machine learning and statistical methods were reviewed theoretically as well as practically in the broader scope of business problem prediction. It was determined that bayesian networks and ensemble tree methods have potential to perform estimation as well as complex algorithms such as neural networks while providing the additional benefit of insights into the reasoning behind model predictions.
